---
title: "Exploratory Data Analysis"
author: "K.D"
date: "2026-04-19"
output: html_document
---

## Setup

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(here)
source(here::here("src", "library.R"))
source(here::here("src", "functions.R"))
source(here::here("src", "output.R"))

knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document highlights some preliminary EDA which uncovers interesting patterns in the margin data. This document focuses on margins as the general properties of GDP and THW are well known.

As a reminder, **GOP margins** are defined as:

$$
\text{GOP Margin} = \frac{\text{Gross Operating Profit}}{\text{Sales}}
$$

while **GOS margins** are defined as:

$$
\text{GOS Margin} = \frac{\text{Gross Operating Surplus}}{\text{Gross Value Added}}
$$

Given that these measures are constructed from different underlying aggregates, they display differing sensitivity to fluctuations in business conditions, resulting in distinct margin dynamics over the business cycle.

## Descriptive statistics

```{r descriptive stats}
desc_stats_ts(gop_wide, value_col = Total)
desc_stats_ts(gos_wide, value_col = Total)
```

`GOP` margins are lower on average (11.85%) than `GOS` margins (45.87%). This may highlight more variability in operating profits or sales over time, as they are more susceptible to changes in economic conditions.

## Distribution Highlights

These plots track profit margins across all selected industries (See Appendix 1.A), particularly highlighting the Total aggregate

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot_highlight(gop, "Total")
# Total sits roughly in the middle of the industry distribution

plot_highlight(gos, "Total") + labs(title = "Gross Operating Surplus Margins")
# Total sits slightly above the median
```

`GOP` margins across sectors are relatively more clustered than `GOS`. This may reflect that there is smaller variation in business profit across sectors compared to the surplus gained from production across sectors. the `GOP` Total aggregate sits just above the median, perhaps being skewed by the significant contributions of the mining sector (the top-most line).

## Variance Decomposition

As we have shown, the margins are skewed by large contributing sectors.

```{r}
c1
c2
```

`GOP` does not account for the agricultural sector, making `Total less mining` (an aggregate subtracting the contributions of the mining sector) the closest comparison measure.

`GOP` becomes significantly reduced and less volatile after abstracting from the mining sector, signifying a more centralised margin measure. The same holds true for `GOS`, as removing mining reduces outlying spikes in the original `Total` series. Further abstraction from agriculture only reduces the level of the series, reflecting that the agricultural sector doesnt drive the shape of the time-series.

This legitimises our approach of using `Total less mining` as a true domestic margin proxy.

## Further Time-Series Analysis

```{r}
gop_ts <- gop |>
  mutate(date = yearquarter(date)) |>
  as_tsibble(key = industry, index = date)

gos_ts <- gos |>
  mutate(date = yearquarter(date)) |>
  as_tsibble(key = industry, index = date)
```

### Seasonal Plots

```{r}
gop_ts |> 
  filter(industry == "Total less mining") |> 
  gg_season(y = margin) + labs(title = "GOP")


gos_ts |> 
  filter(industry == "Total less mining") |> 
  gg_season(y = margin)+ labs(title = "GOS")

```

These seasonal plots visualise changes within a single year, overlaid across multiple years to allow for direct comparison.

`GOP` shows a small increase across time, following the general increase in macroeconomic variables over time. However, there is a significant spike occurring during 2020 (highlighted by the corresponding blue colour), the same holds true for `GOS`, highlighting that the COVID-19 period is a significant outlier for the series.

### Subseries plots

Subseries plots visualise the change in each quarter across the length of the time series, and are valuable to anaylse particular outliers in a series.

```{r}
gop_ts |> 
  filter(industry == "Total less mining") |> 
  gg_subseries()


gos_ts |> 
  filter(industry == "Total less mining") |> 
  gg_subseries()
# The change has been much more steady, with Q2 and Q3 having less of a response to COVID-19
#!! shows gos is less volatile than GOP, possible connection to GOP being more responsive to business cycle (conjecture)

```

`GOP` has a more rigid structure in its changed across time, reflecting higher volatility. There is a break in the series at 2015\~, with a two significant spikes occurring during 2020, with the largest impact happening over Q3 of that year.\
The change is much more steady for `GOS`, with Q2 and Q3 having less of a response to COVID-19.

### STL Decomposition

An STL Decomposition splits a time series into three additive components:

$$Y_t = T_t + S_t + R_t$$

Where: \* $Y_t$ (Observed Data): The raw margin percentage at time $t$. \* $T_t$ (Trend-Cycle): The long-term progression of the series, capturing the structural movement of margins while smoothing out high-frequency noise. \* $S_t$ (Seasonal): The recurring intra-year fluctuations. $R_t$ (Remainder): The residual component representing stochastic "shocks" or noise not captured by the trend or seasonality.

```{r}
gop_ts |>
  filter(industry == "Total less mining") |>
  model(STL(margin)) |>
  components() |>
  autoplot() + labs(title = "GOP - STL Decomposition")

gos_ts |>
  filter(industry == "Total less mining") |>
  model(STL(margin)) |>
  components() |>
  autoplot() + labs(title = "GOS - STL Decomposition")
```

[GOP:]{.underline}

-   **Trend:** The series exhibits a cyclical, wavy pattern that remained relatively stable until a significant spike during the COVID-19 pandemic; it has since reverted to its pre-pandemic trajectory, suggesting potential for AR modeling.

-   **Seasonality:** The data shows severe heteroskedasticity, with seasonal variance expanding well before the pandemic. Because seasonal maxima peaked prior to COVID-19, the pandemic cannot be cited as the sole driver of margin expansion.

-   **Remainder:** Persistent patterns in the residuals indicate that the decomposition failed to capture all underlying variation from the original series.

[GOS]{.underline}:

-   **Overall Dynamics:** The series mirrors the structural dynamics observed in the GOP decomposition.

-   **Seasonality:** Seasonal amplitude tapers toward the middle of the series before expanding in later periods.

Both GOP and GOS margins exhibit cyclical trends with significant COVID-period spikes that have since normalized, alongside persistent residual patterns and heteroskedastic seasonality that expanded well before the pandemic.
